---
title: "Analysis for Project"
output: 
  html_document: 
    fig_height: 6
    fig_width: 9
    toc: yes
    toc_depth: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
source("useful_functions.R")
load("output/analytical_data.RData")
library(ggplot2)
library(texreg)
```

# Introduction
<a href="#top">Back to top</a>

This project is looking at the relationship between female physcians and maternal mortality rates. The female physician data are self reported by the patient. Indiviudals were asked if their usual medical care provider was a male or female. I have figures measuring the relationship between female doctors and maternal mortality rate (mmr) as well as figures looking at the trends of mmr and the percent of female doctors over time. 

```{r name, fig.cap="maternal mortality rate by percent of female doctors", fig.align="center"}
ggplot(mm, aes(x=percent_female, y=mmr))+
  geom_point(aes(color=region))+
  geom_smooth(method="lm", se=FALSE)+
  geom_smooth(method="lm", se=FALSE, aes(color=region, group=region))+
  scale_y_log10()+
  theme_bw()

```

```{r}
ggplot(mm, aes(x=year, y=mmr))+
  geom_point(aes(color=region))+
  geom_smooth(method="lm", se=FALSE)+
  geom_smooth(method="lm", se=FALSE, aes(color=region, group=region))+
    scale_y_log10()+
    theme_bw()
```
```{r}
ggplot(mm, aes(x=year, y=percent_female))+
  geom_point(aes(color=region))+
  geom_smooth(method="lm", se=FALSE)+
  geom_smooth(method="lm", se=FALSE, aes(color=region, group=region))+
    scale_y_log10()+
    theme_bw()
```

```{r models, results='asis'}
model1 <- lm(log(mmr)~percent_female, data=mm)
#summary(model1)
model.fixed_region <- lm(log(mmr)~percent_female+region, data=mm)
#summary(model.fixed_region)
model.fixed_both <- lm(log(mmr)~percent_female+region+as.factor(year), data=mm)
#summary(model.fixed_both)
htmlreg(list(model1, model.fixed_region, model.fixed_both), digits=3)

model2 <- lm(year~log(mmr), data=mm)
model2

model3 <- lm(year~percent_female, data=mm)
model3
```


